# Data Preparation {#sec-data-preparation}

Every `gePoints` workflow begins with a data frame. How that data
frame is built matters — not just for the code, but for the clarity
and reproducibility of the document. This chapter describes two
strategies for bringing data into a `gePoints` analysis, and argues
for a discipline of presenting data in formatted tables from the
start.

## Strategy 1: Inline data

For small datasets — roughly 5 to 30 rows — the best approach is to
embed the data directly in the code chunk using `read_csv()` with an
inline string. The data is visible, the columns are aligned, and the
reader sees exactly what the analysis operates on:

```{r}
#| eval: false
library(readr)
library(gt)

sites <- read_csv(
"text,             lat,        lon
Harvard Forest,    42.53690,  -72.17266
Ordway-Swisher,    29.68927,  -81.99343
Konza Prairie,     39.10077,  -96.56390
Yellowstone,       44.95350, -110.53914
Toolik Lake,       68.66109, -149.37047"
)

gt(sites) |>
  tab_caption(caption = "Selected NEON Core Sites") |>
  tab_source_note(source_note = "Source: NEON Strategy Document, 2011")
```

Aligning the columns takes a few minutes but eliminates an entire
category of errors. A latitude value in the longitude column stands
out immediately when the numbers are stacked. With unaligned CSV,
that same error is invisible.

This is the approach used throughout the example chapters of this
book. The datasets are small — 8 to 21 rows — and the visual
inspection of aligned data is worth the formatting effort.

## Strategy 2: External CSV files

For larger datasets, or data that updates independently of the
document, reading from an external file is appropriate:

```{r}
#| eval: false
sites <- read_csv("data/NEON_core_sites.txt")
```

When using external files, the data is hidden — the reader cannot
see it without opening the file separately. A `gt` table in the
rendered document compensates for this, but the code itself is less
self-contained.

External files also introduce path dependencies. The code chunk must
find the file relative to the document's working directory, which
varies depending on whether you render from the command line, from
RStudio, or from a Quarto project. For reproducible documentation,
inline data avoids this problem entirely.

## The `gt` table discipline

Every data frame in this book is displayed as a formatted `gt` table
with a title and a source note. This is not decorative — it serves
three purposes:

1. **Verification.** The table shows exactly what `create_kml()` will
   receive. If a coordinate is wrong, it is wrong in the table too,
   and a careful reader will catch it.
2. **Provenance.** The source note records where the data came from.
   This is metadata that a bare data frame does not carry.
3. **Presentation.** A well-formatted table in the rendered document
   respects the reader. It signals that the data was inspected, not
   just dumped.

The `gt` package is already a dependency of `gePoints`, so no
additional installation is needed.

## Column naming conventions

`create_kml()` expects specific column names. When your source data
uses different names — `Latitude` instead of `lat`, `Site_Name`
instead of `text` — you have two choices. For inline data, simply
name the columns correctly when you type them. For external files,
rename after import:

```{r}
#| eval: false
library(dplyr)

sites <- read_csv("data/external_file.csv") |>
  rename(
    text = Site_Name,
    lat  = Latitude,
    lon  = Longitude
  )
```

The inline approach is simpler: you control the column names at the
point of entry. This is another advantage of embedding small datasets
directly in the code.

## Adding styling columns

Styling columns — `color`, `symbol`, `symbol_scale`, `text_color`,
`text_scale` — can be included in the inline data or added after
import with `mutate()`. For datasets where every row has the same
style, `mutate()` is cleaner:

```{r}
#| eval: false
sites <- sites |>
  mutate(
    color  = "green",
    symbol = "paddle"
  )
```

For datasets where style varies by row — color by category, size by
value — including the styling columns in the inline data makes the
mapping explicit and inspectable. The example chapters demonstrate
both approaches.

## Previewing maps with `preview_map()`

The `preview_map()` function takes the same data frame as
`create_kml()` and renders an interactive Google Maps satellite view
directly in a Quarto document or the RStudio viewer. It is a preview
tool — a way to see your points on real imagery before taking the KML
file to Google Earth for the full 3D experience.

```{r}
#| eval: false
create_kml(sites, "sites.kml")
preview_map(sites)
```

The map defaults to satellite imagery. Other options are `"roadmap"`,
`"terrain"`, and `"hybrid"`:

```{r}
#| eval: false
preview_map(sites, map_type = "hybrid")
```

Click any circle to see a popup built from the `text` and `comment`
columns. The circle colors match the `color` column in your data
frame. The map auto-zooms to fit all points.

### Google Maps API key

`preview_map()` requires a Google Maps API key. The function reads
the key from the environment variable `GGMAP_GOOGLE_API_KEY`. If you
have not set this up, the steps are:

1. Go to the [Google Cloud Console](https://console.cloud.google.com/)
   and create a project (or use an existing one).
2. Enable the **Maps JavaScript API** for that project.
3. Create an API key under **Credentials**.
4. Add the key to your `.Renviron` file so R finds it automatically:

```{r}
#| eval: false
# Open .Renviron for editing
usethis::edit_r_environ()
```

Add this line to the file:

```
GGMAP_GOOGLE_API_KEY=your-key-here
```

Save, restart R, and verify:

```{r}
#| eval: false
Sys.getenv("GGMAP_GOOGLE_API_KEY")
```

The key is stored on your machine, never in your source code or
published documents. This is the same environment variable used by
the `ggmap` package, so if you already use `ggmap`, the key is
already in place.

You can also pass the key directly if needed:

```{r}
#| eval: false
preview_map(sites, api_key = "your-key-here")
```

### Circle radius

The circle size auto-scales to the geographic extent of your data.
For fine control, set the radius in meters:

```{r}
#| eval: false
# Smaller circles for dense, local data
preview_map(sites, radius = 2000)

# Larger circles for continent-scale data
preview_map(sites, radius = 50000)
```
