# Example: Cultural Geography — Zundert {#sec-zundert}

Vincent van Gogh was born in 1853 in Zundert, a municipality in the
province of North Brabant in the southern Netherlands. The landscape
around Zundert — flat, agricultural, crossed by roads connecting small
hamlets — shaped his early visual vocabulary. The fields, heath, and
village life of this area appear throughout his later work.

This example maps 21 hamlets in and around Zundert, demonstrating
`gePoints` in a cultural geography context where the point data tells
a story about settlement patterns and landscape.

## The data

`Zundert_hamlets.csv` contains hamlet names with coordinates:

```{r}
#| eval: false
library(gePoints)
library(readr)
library(dplyr)

zundert <- read_csv("data/Zundert_hamlets.csv")
zundert
```

The coordinates cluster tightly — all 21 hamlets fall within a
rectangle roughly 15 km by 15 km. This density makes the Google Earth
overlay particularly effective, because at the zoom level needed to see
all the points, the satellite imagery resolves individual fields and
roads.

## Basic KML

```{r}
#| eval: false
zundert_kml <- zundert |>
  rename(text = text, lat = lat, lon = lon)

create_kml(zundert_kml, "zundert_basic.kml")
```

Twenty-one red pushpins on the Dutch countryside. Even this default
view is interesting — the hamlets are not uniformly distributed. Some
cluster along roads; others sit in relative isolation.

## Styled version

Use paddle markers with a muted color to complement the landscape
rather than dominate it:

```{r}
#| eval: false
zundert_styled <- zundert |>
  rename(text = text, lat = lat, lon = lon) |>
  mutate(
    color        = "blue",
    symbol       = "paddle",
    symbol_scale = 1.0,
    text_color   = "white",
    text_scale   = 0.9,
    comment      = paste0("Hamlet: ", text,
                          "\nLat: ", round(lat, 4),
                          "\nLon: ", round(lon, 4))
  )

create_kml(zundert_styled, "zundert_styled.kml")
```

## What the map reveals

Zoom to the level where all 21 markers are visible and the settlement
pattern is clear. The hamlets form a loose network connected by roads
that follow field boundaries. Most sit at intersections or along
waterways. The landscape between them is a patchwork of agricultural
plots — the same fields Van Gogh painted.

Markt marks the center of Zundert proper. Klein Zundert sits just to
the north. The hamlets with directional or descriptive names —
Stuivezand (drift sand), Laarheide (heath), Raamberg — encode
landscape features that are sometimes still visible in the satellite
imagery.

This is not a scientific dataset in the usual sense, but it
demonstrates an important use case for `gePoints`: mapping named
locations from a cultural or historical source onto real terrain. The
technique works equally well for archaeological sites, historical
landmarks, literary geographies, or any collection of named places
with coordinates.

A table of 21 hamlet names tells you very little. The same 21 points
on Google Earth, with satellite imagery at field-scale resolution,
tells you about landscape, distance, connectivity, and the physical
world an artist lived in.
