# Example: Herbaria of the World {#sec-herbaria}

A herbarium is a collection of preserved plant specimens — pressed,
dried, mounted on archival paper, and filed by taxonomic name. The
major herbaria of the world hold millions of specimens, some collected
centuries ago, and serve as the primary reference for plant taxonomy
and biodiversity research. Knowing where the large collections are
matters for any botanist planning a research visit or requesting a
loan.

This example maps the nine largest herbaria, with specimen counts
encoded in the popup balloons and marker scaling proportional to
collection size.

## The data

Two tables are needed: collection information and coordinates. They
share the standard herbarium abbreviation code as a key.

```{r}
#| eval: false
library(gePoints)
library(readr)
library(dplyr)
library(gt)

herbaria <- read_csv(
"Name,                                    Abbr, City,             Size,   Million
Museum National d'Histoire Naturelle,     P,    Paris,            Over,   9.5
Royal Botanic Gardens Kew,                K,    London,           Over,   8
New York Botanical Garden,                NY,   New York City,    Nearly, 8
Missouri Botanical Garden,                MO,   St. Louis,        Nearly, 7
Komarov Botanical Institute,              LE,   St. Petersburg,   About,  6
Conserv. et Jardin botaniques de Geneve,  G,    Geneva,           About,  6
Naturhistorisches Museum Wien,            W,    Vienna,           About,  5.5
The United States National Herbarium,     US,   Washington DC,    Over,   5
Harvard University Herbaria,              HUH,  Cambridge,        Over,   5"
)

gt(herbaria) |>
  tab_caption(caption = "Herbarium Information") |>
  tab_source_note(source_note = "Source: Wikipedia")
```

```{r}
#| eval: false
coords <- read_csv(
"Abbr,  lat,       lon
P,      48.8566,    2.3522
K,      51.4839,   -0.3081
NY,     40.7128,  -74.0060
MO,     38.6270,  -90.1994
LE,     59.9343,   30.3351
G,      46.2044,    6.1432
W,      48.2082,   16.3738
US,     38.9072,  -77.0369
HUH,    42.3770,  -71.1167"
)

gt(coords) |>
  tab_caption(caption = "Herbarium Coordinates") |>
  tab_source_note(source_note = "Source: Anthropic Claude lookup")
```

Join on the abbreviation code:

```{r}
#| eval: false
herb_data <- herbaria |>
  left_join(coords, by = "Abbr")
```

## Basic KML

```{r}
#| eval: false
herb_kml <- herb_data |>
  mutate(
    text    = Abbr,
    comment = paste0(Name, "\n", City, "\n",
                     Size, " ", Million, " million specimens")
  )

create_kml(herb_kml, "herbaria_basic.kml")
```

## Styled version with scaled markers

Larger collections get larger markers. The `Million` column is already
numeric, so the scaling is direct:

```{r}
#| eval: false
herb_styled <- herb_data |>
  mutate(
    text         = Abbr,
    comment      = paste0(Name, "\n", City, "\n",
                          Size, " ", Million, " million specimens"),
    symbol_scale = 0.8 + (Million / 10),
    color        = "green",
    symbol       = "paddle",
    text_color   = "white"
  )

create_kml(herb_styled, "herbaria_styled.kml")
```

The scaling formula `0.8 + (Million / 10)` maps the 5–9.5 million
range to marker sizes of 1.3–1.75. Paris (9.5M) is visibly the
largest; the 5-million institutions are noticeably smaller. No text
parsing needed — the numeric `Million` column handles it cleanly.

## What the map reveals

The nine largest herbaria are concentrated in three regions: Western
Europe (Paris, London, Geneva, Vienna), the northeastern United States
(New York, Cambridge, Washington), and St. Louis. A single point
sits in isolation at St. Petersburg.

The clustering is not accidental. The great herbaria were built during
the age of botanical exploration, when European colonial powers and
wealthy American institutions funded collecting expeditions worldwide.
The geographic concentration of specimens in the global north, drawn
from ecosystems across all continents, is one of the enduring
structural features of botanical science.

Plotting these nine points on Google Earth makes that structure visible
at a glance.
