# Example: Cultural Geography — Zundert {#sec-zundert}

Vincent van Gogh was born in 1853 in Zundert, a municipality in the
province of North Brabant in the southern Netherlands. The landscape
around Zundert — flat, agricultural, crossed by roads connecting small
hamlets — shaped his early visual vocabulary. The fields, heath, and
village life of this area appear throughout his later work.

This example maps 21 hamlets in and around Zundert, demonstrating
`gePoints` in a cultural geography context where the point data tells
a story about settlement patterns and landscape.

## The data

```{r}
#| eval: false
library(gePoints)
library(readr)
library(dplyr)
library(gt)

zundert <- read_csv(
"text,                        lat,       lon
Achtmaal,                     51.45250,  4.58194
Rijsbergen,                   51.51694,  4.69611
Wernhout,                     51.45583,  4.64250
De Berk,                      51.57667,  5.77333
De Lent,                      52.44361,  5.74778
Helpt Elkander,               52.00927,  5.93669
't Kapelleke,                 51.70250,  5.90725
Klein Zundert,                51.47480,  4.64310
Klein-Zundertse Heikant,      51.45900,  4.64570
Laarheide,                    51.45550,  4.70700
Laer-Akkermolen,              51.49500,  4.75150
Markt,                        51.45660,  4.64110
Molenstraat,                  51.47230,  4.64870
Poteind,                      51.48490,  4.61240
Raamberg,                     51.44410,  4.65890
Stuivezand,                   51.43420,  4.60730
't Stuk,                      51.54210,  5.81640
Schijf,                       51.49140,  4.54020
Tiggelaar,                    51.47230,  4.70270
Veldstraat,                   51.47040,  4.61150"
)

gt(zundert) |>
  tab_caption(caption = "Hamlets of Zundert") |>
  tab_source_note(source_note = "Source: Compiled from municipal records")
```

The coordinates cluster tightly — most hamlets fall within a
rectangle roughly 15 km by 15 km. This density makes the Google Earth
overlay particularly effective, because at the zoom level needed to see
all the points, the satellite imagery resolves individual fields and
roads.

## Basic KML

```{r}
#| eval: false
create_kml(zundert, "zundert_basic.kml")
```

Twenty-one red pushpins on the Dutch countryside. Even this default
view is interesting — the hamlets are not uniformly distributed. Some
cluster along roads; others sit in relative isolation.

## Styled version

Use paddle markers with a muted color to complement the landscape
rather than dominate it:

```{r}
#| eval: false
zundert_styled <- zundert |>
  mutate(
    color        = "blue",
    symbol       = "paddle",
    symbol_scale = 1.0,
    text_color   = "white",
    text_scale   = 0.9,
    comment      = paste0("Hamlet: ", text,
                          "\nLat: ", round(lat, 4),
                          "\nLon: ", round(lon, 4))
  )

create_kml(zundert_styled, "zundert_styled.kml")
```

## What the map reveals

Zoom to the level where all 21 markers are visible and the settlement
pattern is clear. The hamlets form a loose network connected by roads
that follow field boundaries. Most sit at intersections or along
waterways. The landscape between them is a patchwork of agricultural
plots — the same fields Van Gogh painted.

Markt marks the center of Zundert proper. Klein Zundert sits just to
the north. The hamlets with directional or descriptive names —
Stuivezand (drift sand), Laarheide (heath), Raamberg — encode
landscape features that are sometimes still visible in the satellite
imagery.

This is not a scientific dataset in the usual sense, but it
demonstrates an important use case for `gePoints`: mapping named
locations from a cultural or historical source onto real terrain. The
technique works equally well for archaeological sites, historical
landmarks, literary geographies, or any collection of named places
with coordinates.

A table of 21 hamlet names tells you very little. The same 21 points
on Google Earth, with satellite imagery at field-scale resolution,
tells you about landscape, distance, connectivity, and the physical
world an artist lived in.
