# Example: Ecological Research Sites {#sec-neon}

The National Ecological Observatory Network (NEON) operates 20 core
research sites spanning the continental US, Alaska, Hawaii, and Puerto
Rico. Each site represents one of 20 eco-climatic domains — from
tundra at Toolik Lake on Alaska's North Slope to tropical forest at
Olaa on the Big Island of Hawaii. These sites form the backbone of a
continental-scale ecological monitoring program designed to run for 30
years.

Plotting the 20 core sites on Google Earth reveals the spatial logic
of NEON's design: the domains tile the continent, and each core site
sits within its domain's characteristic ecosystem. The distances
between sites, and their positions relative to terrain, coastlines,
and climate gradients, are immediately visible in a way that a table
of coordinates cannot convey.

## The data

```{r}
#| eval: false
library(gePoints)
library(readr)
library(dplyr)
library(gt)

neon <- read_csv(
"Domain, Domain_Name,              text,                                      lat,        lon
1,       Northeast,                Harvard Forest,                            42.53690,  -72.17266
2,       Mid-Atlantic,             Smithsonian Conservation Biology Inst,     38.89292,  -78.13950
3,       Southeast,                Ordway-Swisher Biological Station,         29.68927,  -81.99343
4,       Atlantic Neotropical,     Guanica Forest,                            17.96955,  -66.86870
5,       Great Lakes,              UNDERC,                                    46.23388,  -89.53725
6,       Prairie Peninsula,        Konza Prairie Biological Station,          39.10077,  -96.56390
7,       Appalachian/Cumberland,   Oak Ridge,                                 35.96412,  -84.28260
8,       Ozarks Complex,           Talladega National Forest,                 32.95046,  -87.39327
9,       Northern Plains,          Woodworth,                                 47.12802,  -99.24133
10,      Central Plains,           Central Plains Experimental Range,         40.81554, -104.74543
11,      Southern Plains,          Caddo/LBJ National Grassland,              33.40123,  -97.57000
12,      Northern Rockies,         Yellowstone Northern Range,                44.95350, -110.53914
13,      Southern Rockies,         Niwot Ridge/Mountain Research Station,     40.05421, -105.58217
14,      Desert Southwest,         Santa Rita Experimental Range,             31.91068, -110.83549
15,      Great Basin,              Onaqui-Ault,                               40.17759, -112.45244
16,      Pacific Northwest,        Wind River Experimental Forest,            45.82049, -121.95191
17,      Pacific Southwest,        San Joaquin,                               37.10872, -119.73156
18,      Tundra,                   Toolik Lake,                               68.66109, -149.37047
19,      Taiga,                    Caribou Creek - Poker Flats Watershed,     65.15401, -147.50258
20,      Pacific Tropical,         Olaa,                                      19.55479, -155.26418"
)

gt(neon) |>
  tab_caption(caption = "NEON Core Research Sites") |>
  tab_source_note(source_note = "Source: NEON Strategy Document, 2011")
```

## Basic KML

For a first pass, use default styling:

```{r}
#| eval: false
create_kml(neon, "neon_core_sites.kml")
```

Open `neon_core_sites.kml` in Google Earth. Twenty red pushpins span
the continent from Puerto Rico to the Arctic. The spatial coverage is
striking — this is a network designed to capture the full range of
North American ecosystems.

## Styled version

Color-coding by broad region makes the geographic pattern clearer:

```{r}
#| eval: false
neon_styled <- neon |>
  mutate(
    color = case_when(
      Domain <= 4  ~ "green",     # Eastern
      Domain <= 8  ~ "blue",      # Central-east
      Domain <= 11 ~ "yellow",    # Plains
      Domain <= 15 ~ "red",       # Western
      Domain <= 17 ~ "purple",    # Pacific
      TRUE         ~ "lightblue"  # Arctic/Tropical
    ),
    symbol       = "paddle",
    symbol_scale = 1.5,
    comment      = paste0("Domain ", Domain, ": ", Domain_Name)
  )

create_kml(neon_styled, "neon_styled.kml")

preview_map(neon_styled)
```

Now the domain structure is visible: green paddles cluster in the
east, yellow across the plains, red through the mountain west, purple
along the Pacific coast, and light blue at the high-latitude and
tropical extremes. The popup balloon for each site shows its domain
number and name.

## What the map reveals

The eastern domains are geographically smaller — higher ecosystem
diversity packed into less area. The western domains are vast, driven
by topography rather than latitude. Alaska's two sites (Toolik Lake
and Caribou Creek) are separated by hundreds of kilometers of roadless
terrain, and the Hawaiian site sits in isolation in the middle of the
Pacific.

This is a 20-point dataset that takes five minutes to process. The
visual payoff is substantial.
