# Example: Kauai Rainfall Gradient {#sec-kauai}

Kauai has one of the steepest rainfall gradients on Earth. The town of
Waimea, on the dry leeward coast, receives about 530 mm of rain per
year. Thirty kilometers away and 1,570 meters up, the summit of Mt.
Waialeale receives over 10,000 mm — among the highest recorded
anywhere on the planet. The factor-of-twenty difference across such a
short distance is driven by orographic lifting: moisture-laden trade
winds hit the mountain, rise, cool, and release their water on the
windward slopes.

This example maps 15 rainfall stations along the gradient, using
marker scaling to encode rainfall amount and color to distinguish
low, moderate, and extreme precipitation zones.

## The data

`Kauai_rainfall.csv` contains station name, coordinates, elevation in
meters, and mean annual rainfall in millimeters:

```{r}
#| eval: false
library(gePoints)
library(readr)
library(dplyr)

kauai <- read_csv("data/Kauai_rainfall.csv")
kauai
```

The rainfall column spans from 531 mm (Waimea) to 10,004 mm (Mt.
Waialeale). The elevation ranges from 6 m at the coast to 1,570 m at
the summit.

## Basic KML

A first pass with station names as labels:

```{r}
#| eval: false
kauai_basic <- kauai |>
  rename(text = text, lat = lat, lon = lon)

create_kml(kauai_basic, "kauai_basic.kml")
```

In Google Earth, 15 red pushpins appear scattered across Kauai. The
spatial pattern is already informative — stations cluster along the
coast and thin out toward the interior — but the rainfall values are
invisible.

## Styled version: color and scale by rainfall

Encode rainfall in both color and marker size:

```{r}
#| eval: false
kauai_styled <- kauai |>
  mutate(
    comment = paste0(
      "Elevation: ", round(elevation), " m\n",
      "Rainfall: ", format(round(rainfall), big.mark = ","), " mm/yr"
    ),
    color = case_when(
      rainfall < 1000  ~ "yellow",     # Dry
      rainfall < 2000  ~ "green",      # Moderate
      rainfall < 4000  ~ "blue",       # Wet
      TRUE             ~ "purple"      # Extreme
    ),
    symbol = "paddle",
    symbol_scale = 0.8 + (log10(rainfall) / 4),
    text_color = "white",
    text_scale = 0.8
  )

create_kml(kauai_styled, "kauai_rainfall.kml")
```

The log scaling is important. A linear scale from 531 to 10,004 would
make the dry-coast markers nearly invisible. The logarithmic transform
compresses the range so that all stations are legible while still
showing clear size differences between the dry coast and the wet
summit.

## What the map reveals

Open `kauai_rainfall.kml` in Google Earth and the orographic pattern
jumps out. The yellow markers line the southwest coast. Green markers
appear at moderate elevations. Blue markers cluster on the windward
slopes. The purple markers — Mt. Waialeale and its neighbors — sit in
the interior highlands.

The 3D terrain in Google Earth makes the causal mechanism visible in a
way no flat map can. Toggle on terrain view and the relationship
between elevation, aspect, and rainfall is unmistakable. The windward
slopes are green and deeply eroded; the leeward coast is brown and
dry. The markers sit in their landscape context.

Click any marker to see the popup: elevation and rainfall for that
station. The transition from 531 mm at Waimea to 10,004 mm at
Waialeale — a factor of nearly 20 — happens over a horizontal
distance shorter than a half-marathon.

This is a 15-point dataset, but it tells a complete geophysical story.
