# Example: Ecological Research Sites {#sec-neon}

The National Ecological Observatory Network (NEON) operates 20 core
research sites spanning the continental US, Alaska, Hawaii, and Puerto
Rico. Each site represents one of 20 eco-climatic domains — from
tundra at Toolik Lake on Alaska's North Slope to tropical forest at
Olaa on the Big Island of Hawaii. These sites form the backbone of a
continental-scale ecological monitoring program designed to run for 30
years.

Plotting the 20 core sites on Google Earth reveals the spatial logic
of NEON's design: the domains tile the continent, and each core site
sits within its domain's characteristic ecosystem. The distances
between sites, and their positions relative to terrain, coastlines,
and climate gradients, are immediately visible in a way that a table
of coordinates cannot convey.

## The data

The file `NEON_core_sites.txt` contains the 20 core sites with domain
number, domain name, site name, latitude, and longitude. The data
comes from the NEON strategy document published in 2011.

```{r}
#| eval: false
library(gePoints)
library(readr)
library(dplyr)

neon <- read_csv("data/NEON_core_sites.txt", skip = 1,
                 col_names = c("domain", "domain_name",
                               "site_name", "lat", "lon"))
neon
```

The file has a header comment on the first line (the source URL), so
we skip it and provide column names directly.

## Creating the KML

For a first pass, map site names to labels and use default styling:

```{r}
#| eval: false
neon_kml <- neon |>
  mutate(text = site_name)

create_kml(neon_kml, "neon_core_sites.kml")
```

Open `neon_core_sites.kml` in Google Earth. Twenty red pushpins span
the continent from Puerto Rico to the Arctic. The spatial coverage is
striking — this is a network designed to capture the full range of
North American ecosystems.

## Styled version

Color-coding by broad region makes the geographic pattern clearer:

```{r}
#| eval: false
neon_styled <- neon |>
  mutate(
    text = site_name,
    color = case_when(
      domain <= 4  ~ "green",     # Eastern
      domain <= 8  ~ "blue",      # Central-east
      domain <= 11 ~ "yellow",    # Plains
      domain <= 15 ~ "red",       # Western
      domain <= 17 ~ "purple",    # Pacific
      TRUE         ~ "lightblue"  # Arctic/Tropical
    ),
    symbol = "paddle",
    symbol_scale = 1.5,
    comment = paste0("Domain ", domain, ": ", domain_name)
  )

create_kml(neon_styled, "neon_styled.kml")
```

Now the domain structure is visible: green paddles cluster in the
east, yellow across the plains, red through the mountain west, purple
along the Pacific coast, and light blue at the high-latitude and
tropical extremes. The popup balloon for each site shows its domain
number and name.

## What the map reveals

A table of 20 latitude/longitude pairs is difficult to interpret. The
Google Earth overlay makes several things immediately apparent:

The eastern domains are geographically smaller — higher ecosystem
diversity packed into less area. The western domains are vast, driven
by topography rather than latitude. Alaska's two sites (Toolik Lake
and Caribou Creek) are separated by hundreds of kilometers of roadless
terrain, and the Hawaiian site sits in isolation in the middle of the
Pacific.

This is a 20-point dataset that takes five minutes to process. The
visual payoff is substantial.
