# Example: Herbaria of the World {#sec-herbaria}

A herbarium is a collection of preserved plant specimens — pressed,
dried, mounted on archival paper, and filed by taxonomic name. The
major herbaria of the world hold millions of specimens, some collected
centuries ago, and serve as the primary reference for plant taxonomy
and biodiversity research. Knowing where the large collections are
matters for any botanist planning a research visit or requesting a
loan.

This example maps the nine largest herbaria, with specimen counts
encoded in the popup balloons and marker scaling proportional to
collection size.

## The data

Two files are needed. The first, `herbaria.csv`, has names,
abbreviations, cities, and collection sizes:

```{r}
#| eval: false
library(gePoints)
library(readr)
library(dplyr)

herbaria <- read_csv("data/herbaria.csv")
herbaria
```

The second, `herbaria_loc.csv`, has coordinates keyed by the standard
herbarium abbreviation:

```{r}
#| eval: false
coords <- read_csv("data/herbaria_loc.csv")
coords
```

Join them on the abbreviation code:

```{r}
#| eval: false
herb_data <- herbaria |>
  left_join(coords, by = "Abbreviation") |>
  rename(
    lat = Latitude,
    lon = Longitude
  )
herb_data
```

## Creating the KML

The basic version uses herbarium names as labels:

```{r}
#| eval: false
herb_kml <- herb_data |>
  mutate(
    text    = trimws(Name),
    comment = paste0(trimws(City), " — ", trimws(Size), " specimens")
  )

create_kml(herb_kml, "herbaria_basic.kml")
```

## Styled version with scaled markers

Larger collections get larger markers. We parse the size text to
extract approximate numeric values, then scale:

```{r}
#| eval: false
herb_styled <- herb_data |>
  mutate(
    text    = trimws(Abbreviation),
    comment = paste0(trimws(Name), "\n",
                     trimws(City), "\n",
                     trimws(Size), " specimens"),
    # Approximate millions from the Size text
    millions = case_when(
      grepl("9.5", Size) ~ 9.5,
      grepl("8",   Size) ~ 8.0,
      grepl("7",   Size) ~ 7.0,
      grepl("6",   Size) ~ 6.0,
      grepl("5.5", Size) ~ 5.5,
      grepl("5",   Size) ~ 5.0,
      TRUE               ~ 5.0
    ),
    symbol_scale = 0.8 + (millions / 10),
    color  = "green",
    symbol = "paddle",
    text_color = "white"
  )

create_kml(herb_styled, "herbaria_styled.kml")
```

## What the map reveals

The nine largest herbaria are concentrated in three regions: Western
Europe (Paris, London, Geneva, Vienna), the northeastern United States
(New York, Cambridge, Washington), and St. Louis. A single point
sits in isolation at St. Petersburg.

The clustering is not accidental. The great herbaria were built during
the age of botanical exploration, when European colonial powers and
wealthy American institutions funded collecting expeditions worldwide.
The geographic concentration of specimens in the global north, drawn
from ecosystems across all continents, is one of the enduring
structural features of botanical science.

Plotting these nine points on Google Earth makes that structure visible
at a glance. The scaled markers convey relative size — Paris is
clearly the largest, and the five-million-specimen institutions in
Washington and Cambridge are noticeably smaller than Kew or New York.
