# Getting Started

## Installation

`gePoints` is hosted on GitHub. Install it with `devtools`:

```{r}
#| eval: false
# install.packages("devtools")
devtools::install_github("kimbridges/gePoints")
```

The package depends on `xml2` for KML generation, `readr` for data
import, and `rlang` for internal utilities. These will be installed
automatically.

## Your first KML file

The core workflow is three steps: build a data frame, call
`create_kml()`, and open the output in Google Earth.

```{r}
#| eval: false
library(gePoints)
library(readr)
library(gt)

locations <- read_csv(
"text,          lat,       lon
New York City,  40.7484,  -73.9857
Paris,          48.8584,    2.2945
Sydney,        -33.8568,  151.2153"
)

gt(locations) |>
  tab_caption(caption = "Three World Cities") |>
  tab_source_note(source_note = "Coordinates: Google Maps")

create_kml(locations, "three_cities.kml")
```

This produces a file called `three_cities.kml`. Double-click it to
open in Google Earth. You will see three red pushpin markers — one in
Manhattan, one near the Eiffel Tower, one at the Sydney Opera House —
each labeled with the city name.

## What just happened

The `create_kml()` function did several things:

1. Read the data frame and found the required columns: `lat` and `lon`
2. Applied default styling (red pushpin, scale 1.2, white text label)
   to every row because no styling columns were provided
3. Generated the KML XML with one `<Placemark>` element per row
4. Wrote the result to `three_cities.kml`

The only required columns are `lat` and `lon`. Everything else —
labels, colors, marker types, scaling — is optional with sensible
defaults.

## Required and optional columns

| Column | Required | Default | Description |
|--------|----------|---------|-------------|
| `lat` | yes | — | Latitude in decimal degrees |
| `lon` | yes | — | Longitude in decimal degrees |
| `text` | no | — | Label displayed next to the marker |
| `comment` | no | — | Text shown in the popup balloon |
| `color` | no | `"red"` | Marker color |
| `symbol` | no | `"pushpin"` | Marker type: `"pushpin"` or `"paddle"` |
| `symbol_scale` | no | `1.2` | Marker size (0.5–2.0 typical) |
| `text_color` | no | `"white"` | Label text color |
| `text_scale` | no | `1.0` | Label text size |

## Adding a comment column

The `comment` column populates the popup balloon that appears when you
click a marker in Google Earth:

```{r}
#| eval: false
locations <- read_csv(
"text,          lat,       lon,      comment
New York City,  40.7484,  -73.9857, Empire State Building
Paris,          48.8584,    2.2945, Eiffel Tower
Sydney,        -33.8568,  151.2153, Opera House"
)

gt(locations) |>
  tab_caption(caption = "Three Cities with Landmarks") |>
  tab_source_note(source_note = "Coordinates: Google Maps")

create_kml(locations, "cities_with_comments.kml")

preview_map(locations)
```

The `preview_map()` function produces a satellite-view Google Maps
widget showing the same points that `create_kml()` writes to KML.
Click a marker to see its comment in a popup. This is a preview —
the KML file opened in Google Earth provides the full 3D experience.

Now clicking on the Paris marker in Google Earth displays "Eiffel
Tower" in the information balloon. This is useful for attaching
metadata — site descriptions, measurement values, dates — to individual
points without cluttering the map labels.

## Next steps

The next two chapters cover the full range of marker styling: symbol
types and colors (@sec-marker-styles), and text label formatting
(@sec-text-labels). After that, @sec-data-preparation discusses
strategies for organizing your data before calling `create_kml()`.
