# Getting Started

## Installation

`gePoints` is hosted on GitHub. Install it with `devtools`:

```{r}
#| eval: false
# install.packages("devtools")
devtools::install_github("kimbridges/gePoints")
```

The package depends on `xml2` for KML generation, `readr` for data
import, and `rlang` for internal utilities. These will be installed
automatically.

## Your first KML file

The core workflow is three steps: build a data frame, call
`create_kml()`, and open the output in Google Earth.

```{r}
#| eval: false
library(gePoints)

# Three cities, three colors
locations <- data.frame(
  text = c("New York City", "Paris", "Sydney"),
  lat  = c(40.7484, 48.8584, -33.8568),
  lon  = c(-73.9857, 2.2945, 151.2153)
)

create_kml(locations, "three_cities.kml")
```

This produces a file called `three_cities.kml`. Double-click it to
open in Google Earth. You will see three red pushpin markers — one in
Manhattan, one near the Eiffel Tower, one at the Sydney Opera House —
each labeled with the city name.

## What just happened

The `create_kml()` function did several things:

1. Read the data frame and found the required columns: `lat` and `lon`
2. Applied default styling (red pushpin, scale 1.2, white text label)
   to every row because no styling columns were provided
3. Generated the KML XML with one `<Placemark>` element per row
4. Wrote the result to `three_cities.kml`

The only required columns are `lat` and `lon`. Everything else —
labels, colors, marker types, scaling — is optional with sensible
defaults.

## Required and optional columns

| Column | Required | Default | Description |
|--------|----------|---------|-------------|
| `lat` | yes | — | Latitude in decimal degrees |
| `lon` | yes | — | Longitude in decimal degrees |
| `text` | no | — | Label displayed next to the marker |
| `comment` | no | — | Text shown in the popup balloon |
| `color` | no | `"red"` | Marker color |
| `symbol` | no | `"pushpin"` | Marker type: `"pushpin"` or `"paddle"` |
| `symbol_scale` | no | `1.2` | Marker size (0.5–2.0 typical) |
| `text_color` | no | `"white"` | Label text color |
| `text_scale` | no | `1.0` | Label text size |

## Adding a comment column

The `comment` column populates the popup balloon that appears when you
click a marker in Google Earth:

```{r}
#| eval: false
locations <- data.frame(
  text    = c("New York City", "Paris", "Sydney"),
  lat     = c(40.7484, 48.8584, -33.8568),
  lon     = c(-73.9857, 2.2945, 151.2153),
  comment = c("Empire State Building", "Eiffel Tower", "Opera House")
)

create_kml(locations, "cities_with_comments.kml")
```

Now clicking on the Paris marker in Google Earth displays "Eiffel
Tower" in the information balloon. This is useful for attaching
metadata — site descriptions, measurement values, dates — to individual
points without cluttering the map labels.

## Next steps

The next two chapters cover the full range of marker styling: symbol
types and colors (@sec-marker-styles), and text label formatting
(@sec-text-labels). After that, @sec-csv shows how to read data
directly from CSV files, which is the typical workflow for real
datasets.
