# Marker Styles {#sec-marker-styles}

`gePoints` supports two marker types (pushpins and paddles) in up to
nine colors. This chapter shows how to control marker appearance through
the `color`, `symbol`, and `symbol_scale` columns.

## Symbol types

The `symbol` column accepts two values:

- `"pushpin"` — the default; a pin icon that appears to stick into the
  terrain
- `"paddle"` — a circular marker on a stick, useful when you want a
  cleaner look or need to distinguish categories

```{r}
#| eval: false
library(gePoints)

markers <- data.frame(
  text   = c("Pushpin marker", "Paddle marker"),
  lat    = c(35.0, 35.0),
  lon    = c(-110.0, -109.5),
  symbol = c("pushpin", "paddle")
)

create_kml(markers, "symbol_types.kml")
```

## Available colors

Both pushpins and paddles support the following colors:

| Color | Pushpin | Paddle |
|-------|---------|--------|
| red | yes | yes |
| blue | yes | yes |
| green | yes | yes |
| yellow | yes | yes |
| purple | yes | yes |
| white | yes | yes |
| pink | yes | yes |
| lightblue | yes | yes |
| orange | — | yes |

Orange is available only for paddle markers. If you specify orange with
a pushpin, the marker defaults to red.

```{r}
#| eval: false
# Demonstrate all pushpin colors
colors <- c("red", "blue", "green", "yellow",
            "purple", "white", "pink", "lightblue")

color_demo <- data.frame(
  text   = colors,
  lat    = seq(35.0, 35.7, length.out = 8),
  lon    = rep(-110.0, 8),
  color  = colors,
  symbol = rep("pushpin", 8)
)

create_kml(color_demo, "pushpin_colors.kml")
```

## Symbol scaling

The `symbol_scale` column controls marker size. The default is 1.2.
Values between 0.5 and 2.0 cover most practical needs. Smaller values
work well for dense point clouds; larger values make isolated markers
easier to spot at wide zoom levels.

```{r}
#| eval: false
scales <- data.frame(
  text         = c("Small (0.5)", "Default (1.2)", "Large (2.0)"),
  lat          = c(35.0, 35.2, 35.4),
  lon          = c(-110.0, -110.0, -110.0),
  symbol_scale = c(0.5, 1.2, 2.0)
)

create_kml(scales, "symbol_scales.kml")
```

## Mixing styles in one file

Every row in the data frame can have its own styling. This is the
mechanism for encoding categorical variables visually — for example,
color by site type, symbol by data source:

```{r}
#| eval: false
mixed <- data.frame(
  text   = c("Weather station", "Stream gauge",
             "Weather station", "Stream gauge"),
  lat    = c(35.0, 35.1, 35.3, 35.4),
  lon    = c(-110.0, -110.0, -110.5, -110.5),
  color  = c("blue", "green", "blue", "green"),
  symbol = c("pushpin", "paddle", "pushpin", "paddle")
)

create_kml(mixed, "mixed_styles.kml")
```

In Google Earth, the visual distinction between blue pushpins (weather
stations) and green paddles (stream gauges) is immediate. This is a
simple form of thematic mapping — no GIS software required.
