# Text Labels {#sec-text-labels}

Every marker in Google Earth can display a text label — the name that
floats next to the icon on the map. `gePoints` controls labels through
the `text`, `text_color`, and `text_scale` columns.

## The text column

The `text` column provides the visible label. If omitted, markers
appear without labels — useful for dense point clouds where labels
would overlap, though less common in practice.

```{r}
#| eval: false
library(gePoints)

# Labels from the text column
labeled <- data.frame(
  text = c("Summit", "Base camp", "Trailhead"),
  lat  = c(35.4, 35.2, 35.0),
  lon  = c(-111.6, -111.6, -111.6)
)

create_kml(labeled, "labeled_points.kml")
```

## Text color

The `text_color` column sets label color. The default is white, which
reads well against dark terrain and satellite imagery. Available colors
are the same as marker colors (see @sec-marker-styles).

Against light backgrounds — snow, sand, urban areas — white labels
disappear. Switching to black or a dark color solves this:

```{r}
#| eval: false
contrast <- data.frame(
  text       = c("On dark terrain", "On light terrain"),
  lat        = c(35.0, 35.2),
  lon        = c(-111.6, -111.6),
  text_color = c("white", "black")
)

create_kml(contrast, "text_colors.kml")
```

In KML, colors are specified in ABGR (alpha-blue-green-red) hex
format. The `gePoints` package handles this conversion internally
through `get_abgr_color()`, so you work with color names rather than
hex codes.

## Text scaling

The `text_scale` column controls label size. The default is 1.0.
Smaller values (0.6–0.8) reduce clutter when many points are close
together. Larger values (1.2–1.5) make labels legible at wider zoom
levels.

```{r}
#| eval: false
scaling <- data.frame(
  text       = c("Small label", "Normal label", "Large label"),
  lat        = c(35.0, 35.2, 35.4),
  lon        = c(-111.6, -111.6, -111.6),
  text_scale = c(0.6, 1.0, 1.5)
)

create_kml(scaling, "text_scales.kml")
```

## Combining text and marker styles

Text and marker styling are independent. You can set colors, scales,
and symbol types for both markers and labels in the same data frame:

```{r}
#| eval: false
combined <- data.frame(
  text         = c("Primary site", "Secondary site", "Reference"),
  lat          = c(35.0, 35.2, 35.4),
  lon          = c(-111.6, -111.6, -111.6),
  color        = c("red", "blue", "green"),
  symbol       = c("pushpin", "pushpin", "paddle"),
  symbol_scale = c(1.5, 1.2, 1.0),
  text_color   = c("yellow", "white", "green"),
  text_scale   = c(1.2, 1.0, 0.8)
)

create_kml(combined, "full_styling.kml")
```

The resulting KML file encodes site hierarchy visually: the primary
site has a large red pushpin with a bold yellow label, secondary sites
are standard blue pushpins with white text, and reference points are
small green paddles with subdued labels.
