# Working with CSV Data {#sec-csv}

Most real-world use of `gePoints` starts with a CSV file, not a
hand-built data frame. This chapter covers the workflow for reading
external data and preparing it for `create_kml()`.

## Reading a CSV file

The standard approach uses `readr::read_csv()`, which is already a
dependency of `gePoints`:

```{r}
#| eval: false
library(gePoints)
library(readr)

sites <- read_csv("data/SEGA_garden_locations.txt")
sites
```

This produces a tibble with columns `Garden Name`, `Latitude`, and
`Longitude`. The column names don't match what `create_kml()` expects
(`text`, `lat`, `lon`), so we need to rename them.

## Column mapping

`create_kml()` looks for specific column names. When your source data
uses different names, rename before calling the function:

```{r}
#| eval: false
library(dplyr)

sites <- sites |>
  rename(
    text = `Garden Name`,
    lat  = Latitude,
    lon  = Longitude
  )

create_kml(sites, "sega_gardens.kml")
```

This produces a KML file with 8 garden locations in Northern Arizona,
each marked with a default red pushpin.

## Adding styling columns

CSV files rarely include styling information. Add it after import:

```{r}
#| eval: false
sites <- sites |>
  mutate(
    color  = "green",
    symbol = "paddle"
  )

create_kml(sites, "sega_gardens_styled.kml")
```

Now all 8 gardens appear as green paddles. For more complex styling —
color by category, size by value — use standard `dplyr` operations
before the `create_kml()` call:

```{r}
#| eval: false
# Color by latitude band (north vs south)
sites <- sites |>
  mutate(
    color = if_else(lat > 36, "blue", "red")
  )

create_kml(sites, "sega_gardens_by_latitude.kml")
```

## Tab-delimited and other formats

Some of the example datasets use `.txt` extensions with
comma-delimited data. `read_csv()` handles these without issue.
For true tab-delimited files, use `read_tsv()`:

```{r}
#| eval: false
# Comma-delimited .txt file — read_csv works
gardens <- read_csv("data/SEGA_garden_locations.txt")

# Tab-delimited file — use read_tsv
# tab_data <- read_tsv("data/some_tab_file.txt")
```

## Handling messy headers

Real-world CSV files sometimes have extra spaces, inconsistent
capitalization, or multiword column names. The `janitor` package
can help, but for most `gePoints` workflows, explicit `rename()` is
clearest:

```{r}
#| eval: false
# Honolulu tide benchmarks have leading spaces in column names
benchmarks <- read_csv("data/Honolulu_tide_benchmarks.csv")
names(benchmarks)
# [1] "Designation"  " Lat"  " Lon"  " Agency" ...

benchmarks <- benchmarks |>
  rename(
    text = Designation,
    lat  = ` Lat`,
    lon  = ` Lon`
  )

create_kml(benchmarks, "honolulu_benchmarks.kml")
```

The leading spaces in column names are a common CSV artifact. Explicit
renaming makes the mapping visible and reproducible.

## The general pattern

Every CSV-to-KML workflow follows the same steps:

1. `read_csv()` to import the data
2. `rename()` to map columns to `text`, `lat`, `lon`
3. Optionally `mutate()` to add styling columns
4. `create_kml()` to produce the KML file

The example chapters that follow all use this pattern with real
datasets.
